---
output:
  html_document: default
  pdf_document: default
---
##Statistical Inference Course Project 2:  Basic Inferential Data Analysis
Author: E. Mineo
Date: October 7, 2018

In this second project of the course, the R built-in ToothGrowth dataset is used to perform a basic statistical inference exercise.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)
```

1. Exploratory Analysis
```{r}
library(datasets)
library(ggplot2)
d <- ToothGrowth
```

```{r}
str(d)
```

```{r}
summary(d)
```

The dataset contains 60 observations and 3 variables: len, supp and dose. Supp is a factor with two levels "OJ", orange juice and "VC", vitamin C. There are 3 doses: 0.5, 1 and 2mg/day. Len is the response variable. The dataset is plotted in figure 1. The tooth length seems to be different depending on dose and supplement type. Let's verify this using a T-test.


```{r, fig.width=3,fig.height=3}
qplot(dose,len,data=d,ylab = "Length",xlab="Dose VC",color=supp)
```

Fig.1: Plot of the ToothGrowth dataset

2.Comparison of supplement and dose

a) Comparison of supplement
```{r}
g1 <- d[d$supp=="VC",] #Group of animal that received vitamin C
g2 <- d[d$supp=="OJ",] #Group of animal that received orange juice
```

T-Confidence Interval Method
```{r}
n1 <- length(g1$len)
n2 <- length(g2$len)
sp <- sqrt(((n1-1) * sd(g1$len)^2 + (n2-1) * sd(g2$len)^2) / (n1 + n2 - 2))
md <- mean(g2$len) - mean(g1$len)#mean difference
semd <- sp * sqrt(1 / n1 + 1/n2)#standard error of the mean difference
md + c(-1, 1) * qt(.975, n1 + n2 - 2) * semd #T-CI
```

T-test Method
```{r}
t.test(g2$len, g1$len, paired = FALSE)
```
Conclusion: The confidence interval and the p-value show that H0 cannot be rejected with alpha=5%. This means that there is no evidence that the supplement type has an effect on the tooth length.


b) Comparison of doses

Test 1: 0.5 vs. 1 mg/day
```{r}
g3 <- d[d$dose=="0.5",]$len #Group of animals that received 0.5 mg/day
g4 <- d[d$dose=="1",]$len #Group of animals that received 1 mg/day
t.test(g4, g3, paired = FALSE)
```
Conclusion test 1: Hypothesis H0 rejected, there is a significant difference between the effect of the two doses. On average, the length of odontoblasts with a dose of 1 mg/day is greater than with a dose of 0.5 mg/day.


Test 2: 0.5 vs 2 mg/day
```{r}
g5 <- d[d$dose=="2",]$len #Group of animals that received 2 mg/day
t.test(g5, g3, paired = FALSE)
```
Conclusion test 2: Hypothesis H0 rejected, the difference between the effect of the two doses is strongly signififcant (p-value = 4.398e-14). On average, the length of odontoblasts with a dose of 2 mg/day is greater than with a dose of 0.5 mg/day.


Test 3: 1 vs 2 mg/day
```{r}
t.test(g5, g4, paired = FALSE)
```
Conclusion test 3: Hypothesis H0 rejected, there is a significant difference between the effect of the two doses. On average, the length of odontoblasts with a dose of 2 mg/day is greater than with a dose of 1mg/day.

CONCLUSION: Supplement type has no effect on tooth length but dose has a significant effect.